Analyse gender in music psychology from journal publication data

### 2025-09-12 analysis.qmd
Revised figure 2, removed unnecessary plots, added discussion to MS

# Analyse gender in music psychology

T. Eerola, 1/8/2025

Gender analysis of the music psychology journals since 2000 (Musicae Scientiae, Psychology of Music, Music Perception, Journal of New Music Research, and Music & Science).

## Required libraries

@@

## Load and preprocess data

@@

## Quantify

@@

source('scripts/country_table.R') # OK
source('scripts/visualise_gender.R') # OK

@@

```{r}
#| id: keywords
#| eval: true
#| echo: true
#| fig.width: 10
#| fig.height: 6
source('scripts/keyword_analysis.R') 
```

>>>>

---
title: Analyse gender in music psychology from journal publication data
date: "2025-08-01"
author:
  - "Tuomas Eerola"
  - "Anna Czepiel"
format:
  html:
    toc: true
    number-sections: true
    theme: materia
---

Gender analysis of the authors in music psychology journals since 2000. The journals included are _Musicae Scientiae_, _Psychology of Music_, _Music Perception_, _Journal of New Music Research_, and _Music & Science_ and the data been obtained in June 2025 from Scopus.

@@

## Diagnostics

@@

## Quantify authorship types, citations and open access

@@

source('scripts/country_table.R') # OK
#source('scripts/visualise_gender.R') # OK

@@

```{r}
#| id: keywords
#| eval: true
#| echo: true
#| fig.width: 10
#| fig.height: 6
source('scripts/count2category.R') 
source('scripts/keyword_analysis.R') 
```

### 2025-09-12 manuscript/_figure2.qmd
Revised figure 2, removed unnecessary plots, added discussion to MS

```{r}
#| label: fig2
#| fig-width: 9
#| fig-height: 4
#| echo: false
#| warning: false
#| message: false
#| fig.cap: > 
#|   Authorship types across author counts.

@@

custom_palette <- brewer.pal(n = 4, name = "Set1")
custom_palette<-c(custom_palette[4],custom_palette[1:3])
fig2 <- ggplot(author_N,aes(x=author_N_cropped,y=Odds_ratio,color=Type))+
  geom_hline(yintercept=1, linetype="dashed", color = "black") +
  geom_point(position = position_dodge(width = 0.4),shape=15,size=3)+
  geom_errorbar(aes(ymin=CI_low,ymax=CI_high),width=0.2,position = position_dodge(width = 0.4))+
  scale_color_manual(name="Authorship\ntype",values = custom_palette) +
  xlab('Author N')+
  ylab('Female Authorship Odds Ratio')+
  theme_classic(base_size = 15)
print(fig2)
```

>>>>

```{r}
#| label: fig2
#| fig-width: 13
#| fig-height: 4
#| echo: false
#| warning: false
#| message: false
#| fig.cap: > 
#|   A) Number of authors and B) female odds ratio for authorship types across the number of authors.

@@

custom_palette <- brewer.pal(n = 4, name = "Set1")
custom_palette<-c(custom_palette[4],custom_palette[1:3])
fig2 <- ggplot(author_N,aes(x=author_N_cropped,y=Odds_ratio,color=Type))+
  geom_hline(yintercept=1, linetype="dashed", color = "black") +
  geom_point(position = position_dodge(width = 0.4),shape=15,size=3)+
  geom_errorbar(aes(ymin=CI_low,ymax=CI_high),width=0.2,position = position_dodge(width = 0.4))+
  scale_color_manual(name="Authorship\ntype",values = custom_palette) +
  xlab('Number of authors/article')+
  ylab('Female Authorship Odds Ratio')+
  theme_classic(base_size = 15)
#print(fig2)

s <- summarise(group_by(df, BIBTEXKEY), n = n())
S<-s
S$n[S$n > 9]<-10
fig2a <- ggplot(S,aes(x=n))+
  geom_histogram(binwidth=1,fill='grey90',color='black')+
  theme_classic(base_size = 15)+
  labs(x='Number of authors/article',y='Count of articles')+
  scale_x_continuous(breaks=seq(1,10,1),expand = c(0.001,0.001),limits = c(0,11),
                     labels = c(seq(1,9,1),"10+"))+
  scale_y_continuous(expand = c(0.001,0.001))
#print(g)

G<-cowplot::plot_grid(fig2a,fig2,nrow=1,labels=c('A','B'),rel_widths = c(0.4,1))
print(G)

```

### 2025-09-12 manuscript/manuscript.qmd
Revised figure 2, removed unnecessary plots, added discussion to MS

-> to add about topics/keywords:
- @bruck2023bibliometric : in an analysis of medical journals, research keyword varied by gender. For men, it was phase II-III trials, oncology, immunotherapy (etc..), while for mena, it was more related to healthcare-related themes like patient involvement, insurance, quality-of-care, access.
- @sebo2020gender : women more liekly to publish qualitative studies compared to experimental and systematic review articles, perhaps as females demonstrate better relational skill sinvolved in qualitative studies, like prolongeed relationships and emotional ties with research participants. Women were also more likely to publish in primary healthcare journals.
- @ashmos2011gendering : men are under-represented, but women are over-represented in qualitative studies. 
- @thelwall2019gender : females more likely to use exploratory and qualitative rather than quantitative methods. Females more interested in cell biology and veterinary science compared to males, who prefer abstraction, patients, and power/control field
- @vogel2012he : in computer science (computational linguistics), used XX to look at topics that female / male authors published on. Female more likely to publish on dialog, discourse, and sentiment, while men more likely to publish in formal semantic and finite state models.

It is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%). In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067).

To improve certain academic equity issues, it is worthwhile focusing on our discipline and to assess the equity in music psychology. The past efforts have already provided a snapshot a rough overview of the state of affairs (e.g., @anglada2019visualizing) with suggestions of future directions to improve the field. [@eerola2023transparency] showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is, which the view to see how – as a field – we need to diversify participant samples and stimuli to gain a more realistic understanding. In a similar way to these papers, the current paper aims to give a current overview of authorship and gender patterns in the field of music psychology.

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, machine learning, harmony, computation, cognition, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75).

@@

# Discussion
Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several inequities within this aspect of academia, with women often being underrepresented. As such representations differ across different disciplines, it is worthwhile understanding how The goal of the current paper was to provide a snapshot of authorship across gender. While we recognise that gender is a complex and multidimensional social construct, and there are limitations to treating gender as a binary category, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006]. Overall, we see a relatively encouraging pattern within

>>>>

-> to add about topics/keywords/approaches/methods:
- @bruck2023bibliometric : in an analysis of medical journals, research keyword varied by gender. For men, it was phase II-III trials, oncology, immunotherapy (etc..), while for women, it was more related to healthcare-related themes like patient involvement, insurance, quality-of-care, access.
- @sebo2020gender : women more liekly to publish qualitative studies compared to experimental and systematic review articles, perhaps as females demonstrate better relational skill involved in qualitative studies, like prolongeed relationships and emotional ties with research participants. Women were also more likely to publish in primary healthcare journals.
- @ashmos2011gendering : men are under-represented, but women are over-represented in qualitative studies. 
- @thelwall2019gender : females more likely to use exploratory and qualitative rather than quantitative methods. Females more interested in cell biology and veterinary science compared to males, who prefer abstraction, patients, and power/control field
- @vogel2012he : in computer science (computational linguistics), used XX to look at topics that female / male authors published on. Female more likely to publish on dialog, discourse, and sentiment, while men more likely to publish in formal semantic and finite state models.

As the topics, and approaches and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%). In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067).

To improve certain academic equity issues, it is worthwhile focusing on our discipline and to assess the equity in music psychology. The past efforts such as @anglada2019visualizing have already provided a snapshot a rough overview of the state of affairs with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is, which the view to see how – as a field – we need to diversify participant samples and stimuli to gain a more realistic understanding. In a similar way to these papers, the current paper aims to give a current overview of authorship and gender patterns in the field of music psychology.

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, harmony, machine learning, computation, cognition, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75).

@@

# Discussion

Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several inequities within this aspect of academia, with women often being underrepresented. As such representations differ across different disciplines, it is worthwhile to diagnose the situation in music psychology. The goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender. While we recognise that gender is a complex and multidimensional social construct, and there are limitations to treating gender as a binary category, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006].

In music psychology, the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% [@eerola2026inprep]) and higher than in psychology more broadly, where women account for 33.2% of authors [@huang2020]. In terms of authorship roles, notable differences emerge. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. By contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science].

There were substantial gender differences in citation patterns. One notable finding is that first-authored articles by women were significantly more likely to be published open access, with an odds ratio of 1.49, a result that warrants further exploration. This effect does not appear to be tied to specific journals. Although the odds ratio for women publishing open access is highest in _Music and Science_, the broader trend suggests a temporal shift: since around 2010, women have increasingly overtaken men in publishing open access. The data does not provide any reasons for this. While past surveys among UK academics have not indicated substantial differences in experiences with open access [@zhu2017support], financial and other inequalities in emerging countries have been suggested to limit women's capacity to publish in open access journals [@vuong2021adopting].

Turning to potential drivers of these gender differences, geographical and cultural factors appear to play a major role. Our summaries of gender patterns across continents and countries indicate clear cross-national variation. Large Commonwealth nations, namely Australia, Canada, and the United Kingdom, seem to provide a cultural climate more supportive of women in research, as reflected in the relatively high odds ratios of women appearing as authors (in any authorship position) in music psychology. This pattern, however, is not universal across all countries examined [@chan2020]. Prior work shows that women’s research success is positively correlated with national indicators of gender equity [@chan2020]. This is likely to be related to a career length, which strongly correlates with overall productivity. Women exhibit higher dropout rates than men, particularly during the 5–10 years following their first publication [@huang2020], but policies geared towards equity and support for career breaks and caring responsibilities may well contribute towards the observed country-specific differences. Nevertheless, prior analyses demonstrate that even when women constitute a smaller share of researchers within a discipline or country, their contributions are often more impactful than those of their male colleagues [@chan2020]. When examining countries where women have significantly lower odds of authorship compared to men (Japan, Belgium, Austria, Sweden, and Turkey), it is not straightforward to conclude that these nations fall consistently at the lower end of gender equality rankings. The notable exception is Turkey, which is positioned lower according to the Gender Inequality Index [@UNDPHDR2024].

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on care and nurturing [@gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women].
